---
title: "Classical and Bayesian Statistics - Exercises"
toc: true
---

# Problems 1

## a

It seems that, in total, the Allied and Axis countries had about the same number of civilian deaths. However, if we compare the numbers per party member, we see that the Axis countries had a higher average amount of deaths. On the other hand, the data from the Allies is inconsistent (see Denmark).

## b

The percentages in the chart do not sum up to 100%.

## c

### I

We can assume that Linda is a bank teller and is active in the feminist movement. Her experiences from the past could have influenced her behavior and thinking on certain topics. She may have a sense for justice and equality, regardless of the topic.

### II

We can assume that Steve is a librarian because of his helpful personality and his need for order and structure, much like what's found in a library. His passion for details may also be connected to a desire for knowledge.

### III

A Ball costs $0.05.

### IV

There are more death by heart diseases than accidents.

## d

Since the engine is the only thing that keeps a plane in the air, it makes sense for it to be more armoured than the rest of the plane. Even if other parts have more bullet holes on average (e.g. the fuselage), the plane could still fly.

# Problem 1.2

```{r}
# Define vectors
winner <- c(193, 183, 191, 185, 185, 182, 182, 188, 188, 188, 185, 185, 177,
182, 182, 193, 183, 179, 179, 175)

opponent <- c(163, 191, 165, 187, 175, 193, 185, 187, 188, 173, 180, 177, 183,
185, 180, 180, 182, 178, 178, 173)
```

### a

```{r}
# Determine length
cat("Length of vector winners =", length(winner), "\n") # Add line break
cat("Length of vector opponent =", length(opponent))
```

### b

```{r}
cat("Entries 6 to 10 =",winner[6:10]) # Index starts at 1
```

### c

```{r}
cat("Some values from winner:", winner[c(3, 5, 10, 12)]) # Passing a vector for selection
```

### d

```{r}
cat("Current values:",winner[c(8, 9)], "\n")  # Check values
winner[c(8, 9)] <- 189  # Reassign 
cat("New values:", winner[c(8, 9)])  # Check values
```

### e

```{r}
mu_winner <- mean(winner)
mu_opponent <- mean(opponent)

cat("Mean higth of winner vs. opponent:", mu_winner, "vs.", mu_opponent)
```

### f

```{r}
mu_diff <- mu_winner - mu_opponent

cat("Differences between means =", mu_diff)
```

### g

```{r}
var_winner <- var(winner)
sd_winner <- sd(winner)

cat("Variance / Std. deviation of winner:", var_winner, "/", sd_winner)
```

### h

```{r}
my_variance <- function(data){
    mu <- mean(data)
    sum_of_squares <- sum((data - mu)^2)
    variance <- sum_of_squares / (length(data) - 1)
    
    return(variance)
}

my_stdDeviation <- function(variance){
    stdDevition <- sqrt(variance)

    return(stdDevition)
}

my_var_winner <- my_variance(winner)
my_sd_winner <- my_stdDeviation(my_var_winner)

cat("Variance of Winner =", my_var_winner, "\n") # Add line break
cat("Variance of Winner =", my_sd_winner)
```

## Problem 1.3

```{r}
grades <- c(4.2, 2.3, 5.6, 4.5, 4.8, 3.9, 5.9, 2.4, 5.9, 6, 4, 3.7, 5, 5.2, 4.5, 3.6, 5, 6, 2.8, 3.3, 5.5, 4.2, 4.9, 5.1)

grades <- sort(grades) # Sort values

original_mu <- mean(grades)
originla_meadian <- median(grades)

grades[9:11] <- 1 # Reassign values
new_mu <- mean(grades)
new_median <- median(grades)

cat("Original vs. New mean:", original_mu, "vs.", new_mu, "\n") # Add line break
cat("Median stays the same:", originla_meadian, "=", new_median)
```

# Problems 2

## Probmel 2.1

### a

```{r}
# Read data
data <- read.csv('hs25/sa/data/husband_wife.csv')
data
```

### b

```{r}
summary(data)
```

For each column, we see a breef summary qith quantitav and qualitativ informations about the data.

### c

```{r}
age_diff <- data$age.husband - data$age.wife # Calc age difference

boxplot(age_diff)
```

### d

- The median off `age_diff` is located at about 2.5. In middle, the age difference between husbands and wifes are about 2.5cm.
- 50% of the differences are located between about 0 and 5.
- There are more upper than lower outliers. Meaning that there are more more extreme values, wehere the husband his much higher than there wifes.
- Also the values of the upper outliers are bigger than the lower ones.

## Problem 2.2

```{r}
head(InsectSprays) # Preview data from head
```

### a

```{r}
tapply(InsectSprays$count, InsectSprays$spray, mean)
```

### b

```{r}
boxplot(count ~ spray,
        data = InsectSprays
)
```

## Problem 2.3

```{r}
data <- read.csv('hs25/sa/data/Diet.csv')
head(data)
```

```{r}
# Add column weight.los
data$weight.loss <- data$weight6weeks - data$pre.weight
data$weight.loss
```

```{r}
tapply(data$weight.loss, data$Diet, mean)
```

According to the data, the diet 3 seems to have the highest weight lost effect during the 6 week of therapy. Diet 1 and 2 seems to have a more or less same effect in weight lost. Patience with the diet 2 lost in average at less weight during 6 weeks.

```{r}
boxplot(data$weight.loss ~ data$Diet)
```

- Even when diet 3 seems to have the best effect according to the median, the IQR spreads also the most of all three diets.
- Diet 2 spreads over the hole boxplot the most.
- Diet 1 is effectet by some lower outliers.

## Problem 2.4

### a

