---
title: Discrete Response, Time Series and Panel Data - Exercises
author: Nils Rechberger
date: 2026-03-05
---

# Series 1

## Exercise 1.1

What is the expected period (time period of repetition) and the time step for the following timeseries:

- Sunshine duration per month in Basel from 1990 to 2000: Timestep = 1 month, Period = 12 months.
- Number of newborn babies in the city of Zurich per year from 2000 to 2011: Timestep = 1 year, Periode = None.
- Number of reservations in a restaurant for every night during 4 weeks: Timestep = 1 day, Periode = 7 days.
- Water runoff of a river. The data has been collected every day for 4 years: Time step = 1, periode = 365 days.

## Exercise 1.2

```{r}
hstart <- read.table("/home/nils/dev/mscids-notes/fs26/rtp/data/hstart.dat")

ts <- ts(hstart,start=c(1966,1),freq=12)

plot(ts, main="Residential construction in the USA")
```

This is not a stationary time series. A stationary time series is one whose statistical properties (mean, variance, etc) are constant over time. This data fails that definition due to two distinct types of non-stationarity.

- Trend: The long-term "level" of the data shifts upward and then downward.
- Seasonality: There are recurring annual cycles.
- Changing Variance: The intensity of the fluctuations changes over the years.

## Exercises 1.3

### A
```{r}
set.seed(1)

Et <- ts(rnorm(101, 0, 1))

Et[1] <- 0
y1 <- 0 #delete later

for (i in 2:length(Et)) {
    y1[i-1] <- Et[i] - 0.5 * Et[i-1]
}

y1 = y1[2:length(y1)]
ts.y1 = ts(y1)

plot(ts.y1, main="Simulated Time Series Y1", ylab="Y_t", xlab="Time")
abline(h=0, col="red", lty=2)
```

The time series seems to be stationary.

### B

```{r}
set.seed(1)

Et <- ts(rnorm(101, 0, 1))

Et[1] <- 0
y1 <- 0 #delete later

for (i in 2:length(Et)) {
    y1[i] <- y1[i-1] + Et[i]
}

y1 = y1[2:length(y1)]
ts.y1 = ts(y1)

plot(ts.y1, main="Simulated Time Series Y2", ylab="Y_t", xlab="Time")
abline(h=0, col="red", lty=2)
```

The time series seems not to be stationary.

### C

```{r}
set.seed(1)

Et <- ts(rnorm(101, 0, 1))

Et[1] <- 0
y1 <- 0 #delete later

for (i in 2:length(Et)) {
    y1[i] <- 0.5 * y1[i-1] + Et[i]
}

y1 = y1[2:length(y1)]
ts.y1 = ts(y1)

plot(ts.y1, main="Simulated Time Series Y3", ylab="Y_t", xlab="Time")
abline(h=0, col="red", lty=2)
```

The time series seems to be stationary.

### D

```{r}
set.seed(1)

Et <- ts(runif(101, 0.95, 1.05))

Et[1] <- 0
y1 <- 1 #delete later

for (i in 2:length(Et)) {
    y1[i] <- y1[i-1] * Et[i]
}

y1 = y1[2:length(y1)]
ts.y1 = ts(y1)

plot(ts.y1, main="Simulated Time Series Y2", ylab="Y_t", xlab="Time")
abline(h=0, col="red", lty=2)
```

The time series seems not to be stationary.

# Series 2

## Exercise 2.1

### A

```{r}
# create timeseries object
t <- seq(1, 100, length = 100)
data <- 0.5 * t + 1 + runif(100, -1, 1)

ts <- ts(data)
plot(ts)
length(ts)

# Applying backward differencing
ts.2 <- diff(ts, lag = 1)
plot(ts.2)
length(ts.2)
```

The new time series are 99 units long instead of 100.

### B

```{r}
# create timeseries object
t <- seq(1, 100, length = 100)
data <- 2 * t **2 + 3 * t - 1 + runif(100, -200, 200)

ts <- ts(data)
plot(ts)
length(ts)

# Applying backward differencing
ts.2 <- diff(ts, lag = 1)
plot(ts.2)
length(ts.2)

# Again Applying backward differencing
ts.3 <- diff(ts.2, lag = 1)
plot(ts.3)
length(ts.3)
```

Since we applied backward differencing twice, the length of the output time series increased to 98.

## Exercise 2.2

### A

```{r}
hstart <- read.table("/home/nils/dev/mscids-notes/fs26/rtp/data/hstart.dat")

ts <- ts(hstart[, 1], start = c(1966, 1), frequency = 12)
hstart_stl <- stl(ts, s.window = "periodic")

plot(hstart_stl)
```

### B

```{r}
hstart <- read.table("/home/nils/dev/mscids-notes/fs26/rtp/data/hstart.dat")

ts <- ts(hstart[, 1], start = c(1966, 1), frequency = 12)

weights <- c(1, rep(2, 11), 1) / 24
filtered <- filter(ts, filter = weights)

plot(ts)
points(filtered, type = "l", col = "blue")
lines(hstart_stl$time.series[, "trend"], col = "red")
```

### C

```{r}
hstart <- read.table("/home/nils/dev/mscids-notes/fs26/rtp/data/hstart.dat")

ts <- ts(hstart[, 1], start = c(1966, 1), frequency = 12)
plot(ts)

ts.2 <- diff(ts, lag=12) # Using lag 12 bc 12 months per year
plot(ts.2)
```

## Exercise 2.3

```{r}
data(co2)
plot(diff(co2, 1), type = "n", ylim = c(-5, 15))

for (l in 1:12) {
    ts_diff <- diff(co2, lag = l)
    lines(ts_diff, col = rainbow(12)[l])
}
legend("topleft", legend = paste("Lag", 1:12), col = rainbow(12), lty = 1, ncol = 2)
```

# Series 3

## Exercise 3.1

- A -> 2
- B -> 3
- C -> 1
- D -> 4

## Exercise 3.2

```{r}
cbe <- read.table("/home/nils/dev/mscids-notes/fs26/rtp/data/cbe.dat", header = TRUE)
ts <- ts(cbe$elec,  start = c(1958, 1), frequency = 12)
plot(ts)
```

### A 

At this time, it is not advisable to examine the ACF because the time series is not stationary. We can see trends and seasonality over time. We need to clean that up first before analysing the time series.

### B

```{r}
decomp <- decompose(ts, type = "multiplicative")
acf(decomp$random, na.action = na.pass)
```

### C

```{r}
ts.stl.1 <- stl(log(ts), s.window = "periodic")
plot(ts.stl.1$time.series[, "trend"])
decomp <- decompose(ts.stl.1$time.series[, "trend"], type = "multiplicative")
acf(decomp$random, na.action = na.pass)
```

```{r}
monthplot(ts)
ts.stl.2 <- stl(log(ts), s.window = 12)
plot(ts.stl.2$time.series[, "trend"])
decomp <- decompose(ts.stl.2$time.series[, "trend"], type = "multiplicative")
acf(decomp$random, na.action = na.pass)
```

### D

1. With `type = "multiplicative"` we assumes that seasonality is proportional to the trend.
2. We used `stl(log(ts), ...)` because the original time series had an exponential trend.

### E

```{r}
cbe <- read.table("/home/nils/dev/mscids-notes/fs26/rtp/data/cbe.dat", header = TRUE)
ts <- ts(cbe$elec,  start = c(1958, 1), frequency = 12)

ts.2 <- diff(ts, lag=1)
plot(ts.2)
```

```{r}
cbe <- read.table("/home/nils/dev/mscids-notes/fs26/rtp/data/cbe.dat", header = TRUE)
ts <- ts(cbe$elec,  start = c(1958, 1), frequency = 12)

ts.3 <- diff(ts, lag=12)
plot(ts.3)
```

Using differencing alone, we eliminate the trend but do not obtain a fully stationary time series.

# Series 4

## Exercise 4.1

### A

```{r}
#| output: false
arima.sim(list(ar = c(0.8)), n = 1000)
```

### B

```{r}
theo <- ARMAacf(ar = c(0.5, 0.2), lag.max = 100)
sim_data <- arima.sim(model = list(ar = c(0.5, 0.2)), n = 500)
plugin <- acf(sim_data, lag.max = 100, plot = FALSE)$acf

plot(0:100, theo, type = "l")
lines(0:100, plugin)
```

### C

```{r}
plot(0:20, ARMAacf(ar = 0.8, lag.max = 20), type = "h")
```

### D

```{r}
theo_pacf <- ARMAacf(ar = c(0.5, 0.2), lag.max = 100, pacf = TRUE)
sim_data <- arima.sim(model = list(ar = c(0.5, 0.2)), n = 500)
estim_pacf <- pacf(sim_data, lag.max = 100, plot = FALSE)$acf

plot(1:100, theo_pacf, type = "h", ylim = c(-0.1, 0.6))
lines(1:100, estim_pacf, type = "l")
```

## Exercise 4.2

```{r}
data <- read.table("/home/nils/dev/mscids-notes/fs26/rtp/data/kreatin.dat", header=TRUE)
```

### A

The Model should follow a **Random Walk**.

### B

```{r}
ts <- ts(data$gehalt, start=c(1,1))
plot(ts, main="Creatine Concentration of Human Muscular Tissue")
```

```{r}
acf(ts)
```

```{r}
pacf(ts)
```

## Exercise 4.3

```{r}
data <- read.table("/home/nils/dev/mscids-notes/fs26/rtp/data/ts_S3_A2.dat", header=TRUE)
```

### A

```{r}
plot(ts(data$ts1))
abline(h=mean(data$ts1))
```

$ts1$ seems to be stationary.

```{r}
plot(ts(data$ts2))
abline(h=mean(data$ts2))
```

$ts2$ seems not to be stationary.

### B

```{r}
acf(data$ts1)
pacf(data$ts1)
```

The time serie 1 seems to be a $ACF(1)$-process.

```{r}
acf(data$ts2)
pacf(data$ts2)
```

The time serie 1 seems to be a $ACF(2)$-process.

# Series 5

## Task 5.1

### A

#### I

The ACF plot should have a drop at lag 3.

#### II

```{r}
plot(
    0:30,
    ARMAacf(
        ar = c(0.9, -0.5),
        lag.max = 30
    ),
    type = "h",
    ylab = "ACF"
)
```

```{r}
plot(
    1:30,
    ARMAacf(
        ar = c(0.9, -0.5),
        lag.max = 30,
        pacf = TRUE
    ),
    type = "h",
    ylab = "PACF"

)
```

#### III

```{r}
r.sim1 <- arima.sim(
    n = 200,
    model = list(ar = c(0.9, -0.5))
)
```

#### IV

```{r}
ts.1 <- ts(r.sim1)
plot(ts.1)
```

```{r}
acf(ts.1)
```

```{r}
pacf(ts.1)
```

### B

#### I

The PACF plot should have a drop at lag 4.

#### II

```{r}
plot(
    0:30,
    ARMAacf(
        ar = c(0.8, -0.5, -0.4),
        lag.max = 30
    ),
    type = "h",
    ylab = "ACF"
)
```

```{r}
plot(
    1:30,
    ARMAacf(
        ar = c(0.8, -0.5, -0.4),
        lag.max = 30,
        pacf = TRUE
    ),
    type = "h",
    ylab = "PACF"

)
```

#### III

```{r}
r.sim1 <- arima.sim(
    n = 200,
    model = list(ar = c(0.8, -0.5, -0.4))
)
```

#### IV

```{r}
ts.1 <- ts(r.sim1)
plot(ts.1)
```

```{r}
acf(ts.1)
```

```{r}
pacf(ts.1)
```

## Exercise 5.2

### A

#### I

```{r}
roots_i <- polyroot(c(1, -0.5, -2))
abs(roots_i)
```

The process is not stationary.

#### II

```{r}
roots_ii <- polyroot(c(1, -1))
abs(roots_ii)
```

The process is not stationary.

### B

$$
0.5 + x = 1
$$

$$
x = 1 - 0.5
$$

$$
x = 0.5
$$

The process is stationary for $\alpha_2 = 0.5$.

### C

The condition for stationarity is $∣\alpha∣<1$. Therefore, if $∣\alpha∣≥1$, the root $z=\dfrac{1}{\alpha}$​ lies either on the unit circle $(∣z∣=1)$ or inside it $(∣z∣≤1)$. This mathematically confirms that the model is non-stationary in these cases.

# Series 6

## Exercise 6.1

```{r}
yields <- read.table("http://stat.ethz.ch/Teaching/Datasets/WBL/yields.dat", header = FALSE)

t.yields <- ts(yields[, 1])
```

### A

```{r}
acf(t.yields)
```

```{r}
pacf(t.yields)
```

The process could be modelled by a $AR(p)$-Model with the value $p = 3$

### B

```{r}
r.yw <- ar(t.yields, method = "yw", order.max = 3)
r.yw$resid
str(r.yw)
```

Yule-Walker-Koeffizienten:

* $\alpha_1 = -0.32$
* $\alpha_2 = 0.18$

Model:

$$
X_t = -0.32 \cdot X_{t-1} + 0.18 \cdot X_{t-2} + \varepsilon_t
$$

### C

```{r}
r.burg <- ar(t.yields, method = "burg", order.max = 3)
r.burg$resid
str(r.burg)
```

Burg-Koeffizienten:

* $\alpha_1 = -0.332$
* $\alpha_2 = 0.183$

Model:

$$
X_t = -0.332 \cdot X_{t-1} + 0.183 \cdot X_{t-2} + \varepsilon_t
$$

### D

```{r}
r.mle <- ar(t.yields, method = "mle", order.max = 3)
r.mle$resid
str(r.mle)
```

## Exercise 6.2

```{r}
d.force <- read.table("http://stat.ethz.ch/Teaching/Datasets/WBL/kraft.dat", header = FALSE)
ts.force <- ts(d.force[, 1])
```

### A

```{r}
ts.forceA <- window(ts.force, end = 280)
plot(ts.forceA)
```

### B

```{r}
pacf(ts.forceA)
```


```{r}
ar.force <- ar(ts.forceA, method = "ols", order = 6)
ar.force$aic
```

### C

```{r}
ar.force <- arima(ts.forceA, order = c(6, 0, 0), method = "ML")
plot(ar.force$res)
abline(h=0, col="red", lty=2)
```

The model seems to be appropriate for this time series since the residuals looks normal distributed.

# Series 7

## Exercise 7.1

### A

```{r}
arma.1 <-arima.sim(n=250,list(ar=c(-0.75),ma=c(-0.3, 0.25)))
```

### B

```{r}
arma.2 <-arima.sim(n=250,list(ar=c(-0.75, -0.3),ma=c(0.25)))
```

## Exercise 7.2

```{r}
library(fpp2)
data(sunspotarea)
plot(sunspotarea)
```

### B

```{r}
ts.sunspotarea <- ts(log(sunspotarea))
ts.sunspotareaA <- window(ts.sunspotarea, end = 100)
```

```{r}
acf(ts.sunspotareaA)
```

```{r}
pacf(ts.sunspotareaA)
```

Based on the $(p-)acf$, we choose a $AR$-model with the value $p = 3$

## Exercise 7.3

```{r}
t.url <- "http://stat.ethz.ch/Teaching/Datasets/WBL/varve.dat"
d.varve <- ts(scan(t.url)[201:550])
```

### A

```{r}
plot(d.varve)
```

Yes, there is a need for log-transform.

### B

```{r}
plot(log(d.varve))
```

The time series it not stationary by simply using log-tranformation. We try differencing.

```{r}
plot(diff(d.varve, 10))
```

Looks good now.

# Series 8

## Exercise 8.1

### A

```{r}
arima.1 <-arima.sim(
    n=250,
    order=c(2, 1, 2),
    list(
        ar=c(0.5, 0,5),ma=c(-0.4, 0.3)
        )
    )
```

The problem: `ar` part of model is not stationary!

### B

```{r}
# 1. Define the coefficients as polynomials: c(1, -coef1, -coef2)
ar_poly <- c(1, -0.5, -0.5) 
diff_poly <- c(1, -1)       

# 2. Multiply the polynomials using convolution
# We reverse the second vector (diff_poly) because of how convolve() calculates
arma_ar_poly <- convolve(ar_poly, rev(diff_poly), type = "open")

# View the resulting polynomial: 1 - 1.5*B + 0*B^2 + 0.5*B^3
print(arma_ar_poly) 
# Output: [1]  1.0 -1.5  0.0  0.5

# 3. Extract the ARMA coefficients
# Drop the intercept (the first '1') and flip the signs back to standard AR format
arma_ar_coefs <- -arma_ar_poly[-1]

# 4. The MA coefficients remain identical to the original ARIMA model
arma_ma_coefs <- c(-0.4, 0.3)

# Print final coefficients
cat("Equivalent ARMA AR coefficients:", arma_ar_coefs, "\n")
cat("Equivalent ARMA MA coefficients:", arma_ma_coefs, "\n")
```

> Note: Solution generated by AI (Gemini 3.1 Pro).

## Exercise 8.2

```{r}
d.beluga <- read.table("http://stat.ethz.ch/Teaching/Datasets/WBL/beluga.dat", header = TRUE)
d.beluga <- ts(d.beluga)

d.beluga
```

### A

```{r}
y = lm(NURSING ~ ., data=d.beluga)
y
```

```{r}
plot(y$res, type="l")
abline(h=0, col="red")
```

### B

```{r}
acf(d.beluga)
```

```{r}
r.burg <- ar(
    x=d.beluga,
    order.max=8
    )
plot(r.burg$res)
```

### C

```{r}
library(nlme) # Load the package containing the procedure gls()

r.bel.gls <- gls(
    NURSING ~ BOUTS + LOCKONS + DAYNIGHT + PERIOD,
    data = d.beluga,
    correlation = corARMA(
        form = ~PERIOD,
        p = r.burg$order,
        q = 0,
        fixed = FALSE
    ),
    method = "ML"
    )

summary(r.bel.gls)
d.resid <- ts(resid(r.bel.gls))
```

## Exercise 8.3

### A

```{r}
t.url <- "http://stat.ethz.ch/Teaching/Datasets/WBL/varve.dat"
d.varve <- ts(scan(t.url)[201:550])
```

```{r}
acf(diff(log(d.varve)))
```

```{r}
pacf(log(d.varve))
```


```{r}
r.varve <- arima(
    x=log(d.varve),
    order = c(2, 1, 2),
    method="ML"
    )
r.varve
```

### B

$$
y = -0.1442\alpha_1 + 0.1452\alpha_2 + -0.5334\beta_1 + -0.3548\beta_2 + \varepsilon_0
$$

# Series 9

## Exercise 9.1

### A


```{r}
library(astsa)
data(prodn)

plot(prodn)
```

### B

```{r}
plot(diff(log(prodn)))
```

### C

```{r}
prodn.trans <- diff(log(prodn))

acf(prodn.trans)
```

```{r}
pacf(prodn.trans)
```

```{r}
library(forecast)

fit<-arima(
    x=prodn.trans,
    order=c(1,1,1),
    seasonal=c(1,1,1)
    )

tsdisplay(fit$residual)
```

```{r}
fitAuto<-auto.arima(prodn.trans, ic="aic")
fitAuto

tsdisplay(fitAuto$residual)
```
