---
title: Discrete Response, Time Series and Panel Data - Exercises (Rep.)
author: Nils Rechberger
date: 2026-03-05
execute:
  warning: false
  message: false
---

# Series 1

## Exercise 1.1

### a

$12$

### b

None

### c

$7$

### d

$365$

## Exercise 1.2

```{r}
hstart <- read.table("/home/nils/dev/mscids-notes/fs26/rtp/data/hstart.dat")

ts.hstart <- ts(
  data=hstart,
  start=1966,
  end=1974
  )

plot(ts.hstart)
```

The process is not stationary. We see a seasonal trend from 1967 to 1972.

## Exxercise 1.3

### a

```{r}
set.seed(1)

Et <- ts(rnorm(101, 0, 1))
Et[1] <- 0
y1 <- 0 #delete later

for (i in 2:length(Et)) {
y1[i] <- Et[i] - 0.5 * Et[i-1]
}

y1 = y1[2:length(y1)]
ts.y1 = ts(y1)
plot(y1, type="l")
abline(h=mean(y1), col="red")
```

The process seems stationary.

### b

```{r}
set.seed(1)

Et <- ts(rnorm(101, 0, 1))
Et[1] <- 0
y2 <- 0 #delete later

for (i in 2:length(Et)) {
y2[i] <- y2[i-1] + Et[i]
}

y2 = y2[2:length(y2)]
ts.y2 = ts(y2)
plot(y2, type="l")
abline(h=mean(y1), col="red")
```

The process is not stationary.

### c

```{r}
set.seed(1)

Et <- ts(rnorm(101, 0, 1))
Et[1] <- 0
y3 <- 0 #delete later

for (i in 2:length(Et)) {
y3[i] <- 0.5 * y3[i-1] + Et[i]
}

y3 = y3[2:length(y3)]
ts.y3 = ts(y3)
plot(y3, type="l")
abline(h=mean(y3), col="red")
```

The process seems stationary.

### d

```{r}
set.seed(1)

Et <- ts(runif(101, 0.95, 1.05))
Et[1] <- 0
y4 <- 1 #delete later

for (i in 2:length(Et)) {
y4[i] <- y4[i-1] * Et[i]
}

y4 = y4[2:length(y4)]
ts.y4 = ts(y4)
plot(y4, type="l")
abline(h=mean(y4), col="red")
```

The model is not stationary

# Series 2

## Exercise 2.1

### a

```{r}
t <- seq(1, 100, length = 100)
data <- 0.5 * t + 1 + runif(100, -1, 1)
ts <- ts(data)

plot(ts)
print(length(ts))

plot(diff(ts))
print(length(diff(ts)))
```

We can remove the linear trend with one time backward differencing. The length of the ts increases by 1.

### b

```{r}
t <- seq(1, 100, length = 100)
data <- 2 * t**2 + 3 * t - 1 + runif(100, -200, 200)
ts <- ts(data)

plot(ts)
print(length(ts))

plot(diff(ts))
print(length(diff(ts)))

plot(diff(ts, difference=2))
print(length(diff(ts, difference=2)))
```

We can remove the linear trend with two times backward differencing. The length of the ts increases by 2.

## Exercise 2.2

```{r}
hstart <- read.table("/home/nils/dev/mscids-notes/fs26/rtp/data/hstart.dat")

ts.hstart <- ts(
  data = hstart$V1, 
  start = c(1966, 1), # January 1966
  end = c(1974, 12),  # December 1974
  frequency = 12      # Monthly data
)

decomp <- stl(ts.hstart, s.window = "periodic") 
plot(decomp)
```

### b

```{r}
weights <- c(0.5, rep(1, 11), 0.5) / 12
ts.filtertrend <- filter(
  ts,
  filter = weights,
  sides = 2
)

plot(ts.filtertrend)
```

### c

```{r}
hstart <- read.table("/home/nils/dev/mscids-notes/fs26/rtp/data/hstart.dat")

ts.hstart <- ts(
  data = hstart$V1, 
  start = c(1966, 1), # January 1966
  end = c(1974, 12),  # December 1974
  frequency = 12      # Monthly data
)

plot((diff(ts, difference=2)))
```

## Exercise 2.3

```{r}
data(co2)

plot(co2)
acf(co2)
```

# Series 3

## Exercise 3.1

A -> 1
B -> 3
C -> 2
D -> 4

## Exercise 3.2

### a

```{r}
cbe <- read.table("/home/nils/dev/mscids-notes/fs26/rtp/data/cbe.dat", header=TRUE)

ts <- ts(
  data=cbe$elec,
  start=c(1958, 1),
  end=c(1990, 12),
  frequency=12
)

plot(ts)
```

It is not meaningful to interpret the correlogram of this time series because the process is not stationary.

### b

```{r}
decomp <- decompose(ts, type = "multiplicative")
acf(decomp$random, na.action = na.pass)
```

### c

```{r}
elec.stl1 <- stl(log(ts), s.window="periodic")
plot(elec.stl1)

elec.stl2 <- stl(log(ts), s.window=13)
plot(elec.stl2)
```

By simply log-transforming the data does make the process stationary.

# Series 4

## Exercise 4.1

### a

```{r}
data <- arima.sim(list(ar = c(0.8)), n = 1000)
```

### b

```{r}
ARMAacf(ar=c(0.8), lag.max = 10)
```

### c

$p(k) = a^k_1 = 0.8^k$

### d

```{r}
ARMAacf(ar=c(0.8), lag.max = 10)
acf(data)
```

## Exercise 4.2

```{r}
kreatin <- read.table("/home/nils/dev/mscids-notes/fs26/rtp/data/kreatin.dat", header=TRUE)

plot(kreatin$gehalt, type="l")
```

The process should follow white noise.

### b

```{r}
acf(kreatin$gehalt)
```

```{r}
pacf(kreatin$gehalt)
```

White noise fits.

## Exercise 4.3

```{r}
data <- read.table("/home/nils/dev/mscids-notes/fs26/rtp/data/ts_S3_A2.dat", header=TRUE)

ts1 <- data$ts1
ts2 <- data$ts2
```

### a

```{r}
plot(ts1, type="l")
abline(h=mean(ts1), col="red")

plot(ts2, type="l")
abline(h=mean(ts2), col="red")
```

$ts1$ seems stationary, $ts2$ not.

# Series 5

## Exercise 5.1

### a

```{r}
data <- arima.sim(list(ar = c(0.9, -0.5)), n = 100)
plot(data)
```

#### II

```{r}
## Theoretical autocorrelations for X_t = 0.9
## X_t-1 - 0.5 X_t-2 + e_t
plot(0:30, ARMAacf(ar = c(0.9, -0.5), lag.max = 30), type = "h", ylab = "ACF")
## Theoretical partial autocorrelations
plot(1:30, ARMAacf(ar = c(0.9, -0.5), lag.max = 30, pacf = TRUE), type = "h", ylab = "PACF")
```

### III

```{r}
r.sim1 <- arima.sim(n = 200, model = list(ar = c(0.9, -0.5)))
plot(r.sim1)
```

#### IV

```{r}
acf(r.sim1)
```

```{r}
pacf(r.sim1)
```

## Exercise 5.2

### a

#### I

```{r}
polyroot(c(1, 0.5, 2))
```

The process is not stationary

#### II

The process is not stationary since $a_1 = 1$.

### b

```{r}
polyroot(c(1, 0.5, 0.26))
```

$a_2$ has to be $< 0.25$.

# Series 6

## Exercise 6.1

```{r}
yields <- read.table("/home/nils/dev/mscids-notes/fs26/rtp/data/yields.dat", header = FALSE)
t.yields <- ts(yields[, 1])
```

### a

```{r}
acf(t.yields)
```

```{r}
pacf(t.yields)
```

The process can be modeled by a $AR(2)$.

### b

```{r}
r.yw <- ar(t.yields, method = "yw", order.max = 2)
r.yw$resid
str(r.yw)
```

### c

```{r}
r.burg <- ar(t.yields, method = "burg", order.max = 2)
r.burg$resid
str(r.burg)
```

### d

```{r}
r.mle <- ar(t.yields, method = "mle", order.max = 2)
r.mle$resid
str(r.mle)
```

## Exercise 6.2

```{r}
d.force <- read.table("/home/nils/dev/mscids-notes/fs26/rtp/data/kraft.dat", header = FALSE)
ts.force <- ts(d.force[, 1])
```

### a

```{r}
ts.forceA <- window(ts.force, end = 280)
plot(ts.forceA)
```

### b

```{r}
pacf(ts.forceA)
```

We should use a $AR(7)$ model.

```{r}
ar.force <- ar(ts.forceA, method = "ols")
plot(ar.force$aic, type="l")
```

### c

```{r}
ar.force <- arima(ts.forceA, order = c(7, 0, 0), method = "ML")
```
