---
title: Discrete Response, Time Series and Panel Data - Exercise 05
author: Nils Rechberger
date: 2026-05-31
execute:
  warning: false
  message: false
---

# Series 5

## Task 5.1

### A

#### I

The ACF plot should have a drop at lag 3.

#### II

```{r}
plot(
    0:30,
    ARMAacf(
        ar = c(0.9, -0.5),
        lag.max = 30
    ),
    type = "h",
    ylab = "ACF"
)
```

```{r}
plot(
    1:30,
    ARMAacf(
        ar = c(0.9, -0.5),
        lag.max = 30,
        pacf = TRUE
    ),
    type = "h",
    ylab = "PACF"

)
```

#### III

```{r}
r.sim1 <- arima.sim(
    n = 200,
    model = list(ar = c(0.9, -0.5))
)
```

#### IV

```{r}
ts.1 <- ts(r.sim1)
plot(ts.1)
```

```{r}
acf(ts.1)
```

```{r}
pacf(ts.1)
```

### B

#### I

The PACF plot should have a drop at lag 4.

#### II

```{r}
plot(
    0:30,
    ARMAacf(
        ar = c(0.8, -0.5, -0.4),
        lag.max = 30
    ),
    type = "h",
    ylab = "ACF"
)
```

```{r}
plot(
    1:30,
    ARMAacf(
        ar = c(0.8, -0.5, -0.4),
        lag.max = 30,
        pacf = TRUE
    ),
    type = "h",
    ylab = "PACF"

)
```

#### III

```{r}
r.sim1 <- arima.sim(
    n = 200,
    model = list(ar = c(0.8, -0.5, -0.4))
)
```

#### IV

```{r}
ts.1 <- ts(r.sim1)
plot(ts.1)
```

```{r}
acf(ts.1)
```

```{r}
pacf(ts.1)
```

## Exercise 5.2

### A

#### I

```{r}
roots_i <- polyroot(c(1, -0.5, -2))
abs(roots_i)
```

The process is not stationary.

#### II

```{r}
roots_ii <- polyroot(c(1, -1))
abs(roots_ii)
```

The process is not stationary.

### B

$$
0.5 + x = 1
$$

$$
x = 1 - 0.5
$$

$$
x = 0.5
$$

The process is stationary for $\alpha_2 = 0.5$.

### C

The condition for stationarity is $∣\alpha∣<1$. Therefore, if $∣\alpha∣≥1$, the root $z=\dfrac{1}{\alpha}$​ lies either on the unit circle $(∣z∣=1)$ or inside it $(∣z∣≤1)$. This mathematically confirms that the model is non-stationary in these cases.
