---
title: "R-Bootcamp - Notes"
author: "Nils Rechberger"
date: "2026-01"
format: pdf
---

## Intro

### What is R?

R is a powerful programming language and software environment specifically designed for statistical computing, data analysis, and high-quality graphical visualization. It is widely used by researchers and data scientists due to its vast ecosystem of packages and its ability to handle complex data manipulation.

### Steps of a data analysis

1. Import Data 
2. Prepare Data:
3. Graphical Analysis
4. Fit Model
5. Check and interpret results
6. Report results

### R Basics

#### R-Script

Statistical analyses are composed by many steps, where intermediate results are stored, inspected and commented. The sequence of instructions can be collected in a script.

#### Vectors

One-dimensional sequence of elements of a given type (numeric, character (or string), or logical). Can be created with the `c()` function (“c” stands for “combine”).

```{r}
my_vector <- c(1, 2, 3, 4)
my_vector
```

Accessing elements of a vector can be done via indexing, which uses square brackets.

```{r}
v.age <- c(15, 25, 17, 34, 6)
v.age[c(2, 3, 5)] # Several elements accessed simultaneously
```

#### Data Frames

Data often comes in 2-dimensional tables. Data frames are 2-dimension objects. All elements of a given column in a data frame belong to the same class
(e.g. numeric, character, etc.). Data frames can be created via the `data.frame()` function.

#### Lists

Lists are more flexible objects that can store objects of different classes and different dimension (e.g. models). All elements of a list can be accessed using the double squared brackets. Named elements of a list can be accessed via the `$` symbol.

#### Importing Data

Data sets can be imported into R from:

- An existing file
- An internet connection (url)
- A data base (not discussed here)
- etc.

#### Working Directory

R is working/pointing to a specific directory on your machine.

```{r}
getwd() # Check where your current working directory is
```

The working directory can be changed with the `setwd()` function.

## Data Manipulation using dplyr

```{r}
#| echo: false
require(dplyr)
require(tidyr)
data(diamonds, package = "ggplot2")
```

### Long and wide format

- **Wide**: One row per subject, columns for repeated measures.
- **Long**: One row per measurement, with ID and "variable name" columns.

The function `pivot_longer` converts a data frame from wide to long format.

```{r}
diamonds$ID <- 1:nrow(diamonds)
diamonds.long.format <- pivot_longer(
    diamonds, cols = -c(ID, cut, color, clarity),
    names_to = "feature",
    values_to = "value"
)
```

The function `pivot_wider` converts a data frame from long to wide format.

```{r}
pivot_wider(
    diamonds.long.format,
    names_from = feature,
    values_from = value
)
```

> Note: `pivot_longer()` and `pivot_wider()` are function from `tidyr`

### Function of dplyr

#### `select()`

Selecting specific columns in a dataset.

```{r}
select(diamonds, x, y, z) # Select specific cols
```

```r
select(diamonds, -ID) # All cols except "ID"
select(diamonds, contains("ce")) # Cols whose name contains "ce"
select(diamonds, matches("c.")) # Selecting using regex 
```

#### `filter()`

We can use the following function to subset a data frame. It will retain all rows that satisfy the specified condition.

```{r}
filter(diamonds, `carat` > 4, `cut` == "Fair")
```

> Note: The comma `,` between the statements in the function `filter()` means **AND**.

#### `mutate()` and `across()`

Create or calculate new columns.

```{r}
mutate(diamonds, xPLUSy = x + y, depth.check = 2 * z / (xPLUSy))
```

> Note: We can use the newly created `xPLUSy` to calculate `depth.check`.

By combining `mutate()` with `across()`, we can create or calculate new columns for each of the variables/columns.

```{r}
mutate(
    diamonds,
    across(
        -c(cut, color, clarity),
        .fns = cumsum
    )
)
```

#### `summarise()`

Summarise or aggregate.

```{r}
summarise(
    diamonds,
    mean.carat = mean(carat),
    mean.price = mean(price),
    median.carat = median(carat),
    median.price = median(price)
)
```

Again we can apply the `across()` function to summarise or aggregate each of the variables/columns.

```{r}
summarise(
    diamonds,
    across(
        -c(cut, color, clarity),
        .fns = list(
            mean = ~ mean(.x, na.rm = TRUE),
            sum = ~ sum(.x, na.rm = TRUE)
        )
    )
)
```

#### `group_by()`

Group the data such that all functions of `dplyr` are applied to each group of the data separately

```{r}
group_by(
    diamonds,
    cut
)
```

> Note: If you don’t use the `ungroup()` function then the rows of the data frame are still grouped.

### The pipe operator `%>%`

The pipe operator `%>%` makes it easy to read code. Read from left to right, like the natural functions call order.

```{r}
diamonds %>%
    select(x, y, z)
```

> Note: In base R, there is already a pipe operator: `|>`

### Join or Combine Data Sets

#### `left_join()`

A `left_join` returns all observations from `x` and from `y` which could be matched to `x`.

```r
left_join(
    x = data.A,
    y = data.B,
    by = "name"
)
```

#### `right_join()`

A `right_join` returns all observations from `y` and from `x` which could be matched to `y`.

```r
right_join(
    x = data.A,
    y = data.B,
    by = "name"
)
```

#### `inner_join()`

An `inner_join` returns all observations which could be matched in both data sets.

```r
inner_join(
    x = data.A,
    y = data.B,
    by = "name"
)
```

#### `full_join()`

A `full_join` returns all observations from `x` and `y`.

```r
full_join(
    x = data.A,
    y = data.B,
    by = "name"
)
```

::: {.callout-note}
Sometimes, variables are named differently in the data sets that we wish to join. We can still join these data sets as long as we indicate what matches what.

```r
full_join(
    x = data.A,
    y = data.B,
    by = c("name.A" = "name.B")
)
```
:::

## ggplot2
```{r}
#| echo: false
library(ggplot2)
```

`ggplot2` is a powerful and widely-used R package for data visualization based on the Grammar of Graphics, allowing users to create complex plots by layering components like scales, layers, and aesthetics. It excels at handling data in Long format, as shown in your table, making it easy to map variables to visual properties like color, size, and shape.

### Initial Function Call

`ggplot()` function initializes a ggplot object (or conceptually a plot), which can then be modified to create a plot using `ggplot2`.

```r
ggplot(
    data,
    mapping = aes(
        x = x_var,
        y = y_var
    )
)
```

Afterwards, you add layers with the “+” symbol.

```r
ggplot(
    data,
    mapping = aes(
        x = x_var,
        y = y_var
    ) +
) + geom_point()
```

#### Density Plot

You can create density plots. In this case the y argument is not needed.

```{r}
ggplot(
    data = Loblolly,
    mapping = aes(
        x = height
    )
) + geom_density()
```

It is possible to add a "rug" to the plot, showing the value taken by each observation.

```r
ggplot(
    data = Loblolly,
    mapping = aes(
        x = height
    )
) + 
geom_density() +
geom_rug()
```

#### Histogram

Instead of a density plot, you can create histograms.

```{r}
ggplot(
    data = Loblolly,
    mapping = aes(
        x = height
    )
) +
geom_histogram()
```

#### Scatterplot

To create a scatter plot, you need to define x and y, and then the points layer.

```{r}
ggplot(
    data = Loblolly,
    mapping = aes(
        x = age,
        y = height
    )
) +
geom_point()
```

Jitter can be used if over-plotting is a problem. 

> Note: We are log-transforming the x-axis.

```{r}
ggplot(
    data = Loblolly,
    mapping = aes(
        x = age,
        y = height
    )
) +
geom_jitter(
    alpha = 0.2,
    width = 0.5,
    height = 0
) +
scale_y_log10()
```

#### Facetting

Use facetting or panelling to create multiple plots that display different subsets of your data for clearer comparisons.

```{r}
ggplot(
    data = Loblolly,
    mapping = aes(
        y = height,
        x = age
    )
) +
geom_point() +
facet_wrap(~ Seed)
```

#### Colour

Points can be coloured **according** to a certain variable.

```{r}
ggplot(
    data = Loblolly,
    mapping = aes(
        y = height,
        x = age,
        colour = Seed
    )
) +
geom_point()
```

#### Lines

Lines can be drawn by connecting points for each level of a variable by using the "group" argument.

```{r}
ggplot(
    data = Loblolly,
    mapping = aes(
        y = height,
        x = age,
        group = Seed
    )
) +
geom_line(alpha = 0.5)
```

#### Smoothing Methods: Linear Regression

Smoothing line can reveal trends in your data.

```{r}
ggplot(
    data = Loblolly,
    mapping = aes(
        y = height,
        x = age
    )
) +
geom_point() +
geom_smooth(method="lm")
```

::: {.callout-important}
The order of the levels is important. If we swap the last two layers of the previous graph, the points become partly obscured by the regression lines drawn on top.
:::

#### Save a Plot

```r
ggsave(
    filename = "Loblolly_6.png", # Path
    plot = gg.tmp # Graph object
)
```